Praise the people or praise the place?

### 2024-04-23 index.qmd
change slide transition

---
title: "Praise the people or praise the place? Upper tail human capital in electrifying Sweden"
subtitle: "Midway Seminar - 24-01-2024"
author: 
  - "Jonathan Jayes  - Lund University"
title-slide-attributes:
  data-background-image: "assets/preview_wide.png"
  data-background-position: "bottom 10"
  data-background-size: "90%"
  data-background-opacity: "0.4"
bibliography: bibliography.bib
format: 
  revealjs:
    code-overflow: wrap
    width: 1600
    height: 900
    fig-align: center
    default-image-extension: svg
    footer: "[{{< fa house >}}](https://interludeone.com/content/research.html)"
    slide-number: c
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    include-in-header: 
        - text: <link rel = "shortcut icon" href = "assets/favicon.ico" />

@@

How can we use new kinds of source material and new methods to find out where the skilled workers who built the Swedish electricity sector came from, and where they were trained?

Answering involves:

- 2 data structuring tasks
- 2 classification tasks

@@

@@

- 9 of the 11 volumes are digitized by librarians in Uppsala – thank you <3

@@

@@

# Where were the `firms` in which these individuals worked?{#sec-where background-color="#1C9E77" visibility="uncounted"}

## Swedish industrial firm catalogue of 1947

::: {.columns}
::: {.column}
![](assets/svensk-industrikalender_1.png){fig-align="center" width=100%}
:::
::: {.column}
The Svensk Industrikalender, published by Sveriges Industriförbund (Sweden’s Industrial Association), is a comprehensive directory of Swedish industrial firms. This calendar was issued annually from 1918 to 2000 and contains information related to Swedish industry. The 1947 edition available on the Project Runeberg website was digitized in April 2012, sourced from the Centrum för Näringslivshis- toria.
:::
:::

@@

@@

75,000 biographies in 9 volumes of *Vem är Vem?* and 2,400 firms in *Svensk Industrikalender* in a machine readable format for ~ 80 USD.

@@

## Two-part classification process

@@

```{python, eval=FALSE, echo=TRUE}
{
  "input": [
    {
      "occupational_title": "Civilingenjör, teknisk fysik",
      "workplace": "Strömsnäs Järnverks A-B"
    },
  ],
  "output": [
    {x
      "hisco_code": "22",
      "industry": "Iron and steel production"
    }
}

@@

## Results: How far do engineers move?

![](assets/eee_map.png){fig-align="center" width=100%}

# Engineers in electrical appliances and machinery move the furthest

@@

# Conclusion{#sec-conclusion background-color="#1C9E77" visibility="uncounted"}

@@

>>>>

---
title: "Conductors of Progress: Upper Tail Human Capital in Electrifying Sweden"
subtitle: "Presentation at CAS - Senter for grunnforskning"
date: today
author: 
  - "Jonathan Jayes  - Lund University"
title-slide-attributes:
  data-background-image: "assets/preview_wide.png"
  data-background-position: "bottom 10"
  data-background-size: "90%"
  data-background-opacity: "0.4"
bibliography: bibliography.bib
format: 
  revealjs:
    transition: slide
    background-transition: fade
    transition-speed: fast
    code-overflow: wrap
    width: 1600
    height: 900
    fig-align: center
    default-image-extension: svg
    footer: "[{{< fa house >}}](https://interludeone.com/content/research.html)"
    slide-number: c
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    include-in-header: 
        - text: <link rel = "shortcut icon" href = "assets/favicon.ico" />
favicon: "assets/favicon.ico"

@@

@@

## Who were the early electrifiers in Sweden?{visibility="uncounted" transition="none"}

![](assets/map_big_power_stations_line_1900.jpeg){fig-align="center"}

## Who were the early electrifiers in Sweden?{visibility="uncounted" transition="none"}

![](assets/map_big_power_stations_line_1911.jpeg){fig-align="center"}

## Who were the early electrifiers in Sweden?{visibility="uncounted" transition="none"}

![](assets/map_big_power_stations_line_1926.jpeg){fig-align="center"}

## Who were the early electrifiers in Sweden?{#sec-tcg  transition="none"}

![](assets/map_big_power_stations_line_1926_with_treated_and_control_parishes.jpeg){fig-align="center"}

## What were the new jobs due to electrification?

![](assets/06-job_share_plot.jpeg){fig-align="center"}

Who were the highly educated individuals that were at the forefront of the electrification of Sweden and how did they move to opportunity?

## What I want you to take away from today as researchers

::: {.column width="40%"}
1. NLP tasks that used to be very challenging / time consuming are no longer!

2. Classifying occupations is now easier than ever*

::: {.column width="10%"}
:::

::: {.column width="50%"}
![Feel like a world champion](assets/world_champtions.jpeg){fig-align="center" width=80%}
:::

@@

- 8 of the 10 volumes are digitized by librarians in Uppsala – thank you <3

@@

![](assets/vem_cover.png){fig-align="center" width=100%}

@@

## Vem är Vem? Volumes

![](assets/who-is-who-volumes.png){fig-align="center" width=100%}

```{r}
#| eval: false
library(tidyverse)
theme_set(theme_light())
library(ggtext)
library(showtext)
font_add_google("IBM Plex Mono", "ibm")
showtext_auto(enable = TRUE)

books <- books %>%
    mutate(name = str_c(book_name, " - ", year),
            name = fct_reorder(name, pages),
            records = pages * 8) %>%
            mutate(volume = case_when(
                volume == 1 ~ "1945–1950", 
                volume == 2 ~"1962–1968",
                TRUE ~ "Who is who in indsutry and business?")
            ) %>%
    select(name, records, volume)

# Create the ggplot visualization
ggplot(books, aes(x = records, y = name, fill = factor(volume))) +
  geom_col(alpha = .8) +
  labs(x = "Number of biographies", y = "Book volume",
  title = "Number of biographies in Vem är Vem?",
  fill = "Edition") +
  scale_x_continuous(labels = scales::number_format()) +
  scale_fill_manual(values = c("#2B8CBE", "#be5d2b", "midnightblue"))  +
    theme(plot.title.position = "plot") +
    theme(legend.position = "bottom") +
    theme(text = element_text(family = "ibm", size = 24))

ggsave("assets/who-is-who-volumes.png", width = 6, height = 4, units = "in", dpi = 300)
```

## Vem är Vem? Example page

## Vem är Karl Lund?

[Lund, Karl Gustaf]{style="background-color: #b0c4de"}, [chief engineer]{style="background-color: #d2b48c"}, [born on July 22, 1893]{style="background-color: #ff8c69"}, [in Hille, Gävle County]{style="background-color: #90ee90"}, Sweden, son of [clerk Ferdinand L. and Maria Andersson]{style="background-color: #9370db"}. Married in 1936 to [Sigrid Johansson]{style="background-color: #ffadc1"}. Children: [Ingvar (born 1938)]{style="background-color: #ffdead"}, [Lennart (born 1942)]{style="background-color: #ffdead"}. — Graduated from [Bergsskolan]{style="background-color: #afeeee"} in Filipstad in 1917, specialized studies at the [Royal Institute of Technology (KTH)]{style="background-color: #afeeee"} from 1920 to 1922, studied at the [Institute of Metallurgy]{style="background-color: #afeeee"} and [Stockholm University]{style="background-color: #afeeee"} in 1921-1922. Chemist at [Strömsnäs Järnverks A-B]{style="background-color: #98fb98"} in Degerfors from 1918 to 1920, metallurgist and chemist at [Westinghouse Electric & Manufacturing Co.]{style="background-color: #98fb98"} in East Pittsburgh, PA, USA, from 1923 to 1926 and 1928 to 1929, chief metallurgist at [Laclede Steel Co.]{style="background-color: #98fb98"} in Alton, Illinois, USA, in 1927, furnace and steelworks engineer at [A-B Iggesunds Bruk]{style="background-color: #98fb98"} from 1929 to 1931, site manager at [Gunnebo Bruks Nya A-B, Varbergsverket]{style="background-color: #98fb98"}, since 1931. Member of the municipal executive committee, deputy chairman of the economic department, deputy member of the board of the power plant, chairman of [Varbergs Sparbank]{style="background-color: #ffdab9"}, employer representative in the district council of the county labor board, member of the board of [Varbergs Luftskyddsförening (Varberg Air Protection Association)]{style="background-color: #ffdab9"}, secretary of [Varbergs Högerförening (Varberg Conservative Association)]{style="background-color: #ffdab9"}, chairman of the railway sick fund, and [Plant Society for Small Bird Friends]{style="background-color: #ffdab9"}.

## Vem är Karl Lund?

Traveled to [Germany]{style="background-color: #e6e6fa"} in 1921, 1923, 1930, and 1936, [Denmark]{style="background-color: #e6e6fa"}, [Czechoslovakia]{style="background-color: #e6e6fa"} in 1921, 1922, and 1923, [Austria]{style="background-color: #e6e6fa"} in 1921, and the [USA]{style="background-color: #e6e6fa"} from 1923 to 1929. Writings: "Some fundamental factors for obtaining sharp thermal curves" (Transactions of the American Society for Steel Treating, co-authored with C. Benedicks and W. H. Dearden, 1925), "Contemporary production of saw blades, circular saws, and machine knives" (Timber Industry, 1931). Hobbies: [hunting]{style="background-color: #ffd700"} and [fishing]{style="background-color: #ffd700"}.

##

![](assets/tidy.jpg)

@@

## NLP challenges

@@

20,000 biographies in 9 volumes of *Vem är Vem?* and 2,400 firms in *Svensk Industrikalender* in a machine readable format for ~ 80 USD.

## Output

```{python, eval=FALSE, echo=TRUE}
#| code-line-numbers: "6-14"

{
    "original": "Lund, Karl Gustaf, överingenjör, Varberg, f. 22/7/93 i Hille, Gävleb. 1., av brukstj :m. Ferdinand L. o. Maria Andersson. G. 36 m. Sigrid Johansson. Barn: Ingvar f. 38, Lennart 42. — Ex. v. bergssk. i Filipstad 17, spec:-stud. v. KTH (B) 20-22, stud. v. me-tallogr. inst. o. Sthlms högsk. 21-22. Kemist v. Strömsnäs Järnverks A-B, Degerfors, 18-20, metallurg o. kemist v. Westinghouse Electric & Manuf. Co., East Pittsburgh, Pa, USA, 23-26 o. 28-29, chefsmetallurg v. Laclede Steel^Co., Ahon, 111., USA, 27, hytt-o. stålv.ing. v. A-B Iggesunds Bruk 29-31, platschef v. Gunnebo Bruks Nya A-B, Varbergsverket, sed. 31. Led. av drätselkamm., v. ordf. v. ekonomi-avd., suppl. i styr. f. elverket, huv:-man i Varbergs Sparbank, arb :giv. repr. i länsarb:ndns kretsråd, led. av styr. f. Varbergs luftsk ifören., sekr. i Varbergs högerfören., ordfs i järnv. sjukkassa o. Plant :sällsk. Småfågl. Vänner. Res. t. Tyskl. 21, 2^ 23, 30 o. 36, Danm., Tjeckoslov. 21, 22, 23, Österr. 21, USA 23-29. Skr.: Some fundamental factors for obtaining sharp thermal curves (Träns. Am. Soc. for Steel Treating, tills. m. C. Benedicks o. W. H. Dearden 25), Nutida fabrikation av sågblad, sågklingor o. maskinknivar (Trävaruind. 31). Hob-bies: jakt o. fiske.",
    "translated": "Lund, Karl Gustaf, chief engineer, Varberg, born on July 22, 1893, in Hille, Gävle County, Sweden, son of factory worker Ferdinand L. and Maria Andersson. Married in 1936 to Sigrid Johansson. Children: Ingvar (born 1938), Lennart (born 1942). — Graduated from Bergsskolan in Filipstad in 1917, specialized studies at the Royal Institute of Technology (KTH) from 1920 to 1922, studied at the Institute of Metallurgy and Stockholm University in 1921-1922. Chemist at Strömsnäs Järnverks A-B in Degerfors from 1918 to 1920.",
    "structured": {
        "@context": "https://schema.org",
        "@type": "Person",
        "name": "Karl Gustaf Lund",
        "birthDate": "1893-07-22",
        "birthPlace": {
            "name": "Hille, Gävle County, Sweden",
            "latitude": 59.916667,
            "longitude": 15.0
        },
        "jobTitle": "Chief Engineer",
        "memberOf": {
            "@type": "Organization",
            "name": "Swedish Technical Association"
        },
        "children": [
            {
                "@type": "Person",
                "name": "Anita",
                "birthDate": "1937"
            },
            {
                "@type": "Person",
                "name": "Peter",
                "birthDate": "1942"
            }
        ],
        "spouse": {
            "@type": "Person",
            "name": "Marianne Hammarberg",
            "marriageDate": "1936"
        },
        "education": [
            {
                "@type": "EducationalOccupationalCredential",
                "credentialCategory": "Vocational",
                "issuer": {
                    "@type": "Organization",
                    "name": "Technical Gymnasium in Örebro"
                },
                "endDate": "1928"
            },
            {
                "@type": "EducationalOccupationalCredential",
                "credentialCategory": "Degree",
                "issuer": {
                    "@type": "Organization",
                    "name": "KTH"
                },
                "endDate": "1934"
            }
        ],
        "worksFor": {
            "@type": "Organization",
            "name": "Telegraph station in Norrköping",
            "startDate": "1946"
        },
        "hasOccupation": [
            {
                "@type": "Occupation",
                "name": "Engineering assistant",
                "employer": {
                    "@type": "Organization",
                    "name": "Telegraph station in Nässjö"
                },
                "startDate": "1935",
                "endDate": "1935"
            },
            {
                "@type": "Occupation",
                "name": "Line engineer",
                "employer": {
                    "@type": "Organization",
                    "name": "Telegraph station in Norrköping"
                },
                "startDate": "1946"
            }
        ]
    }
}

@@

## Two-part classification process for engineers

@@

```{python, eval=FALSE, echo=TRUE}
{
  "input": [
    {
      "occupational_title": "Civilingenjör, teknisk fysik",
      "workplace": "Strömsnäs Järnverks A-B"
    },
  ],
  "output": [
    {
      "hisco_code": "22",
      "industry": "Iron and steel production"
    }
}

@@

## Assign HISCOs: codes projected from 1536 dimensions to 2 with UMAP

## Results: What are the most common occupations?

![](assets/hisco_counts.png){fig-align="center" width=100%}

# Where were the `individuals` from?{#sec-where background-color="#1C9E77" visibility="uncounted"}

## Where do they come from? Norway edition

```{r}
library(mapview)
library(tidyverse)

df_sf <- read_rds("assets/individual_occupations_institution_geocoded_distances.rds")

df_sf = df_sf %>% 
    mutate(
        narrow_occupation = case_when(
            str_detect(hisco_occupation, "Engineer") ~ "Engineer",
            str_detect(hisco_occupation, "Doctor") ~ "Doctor",
            str_detect(hisco_occupation, "General Managers") ~ "General Manager",
            str_detect(hisco_occupation, "Ministers") ~ "Minister",
            str_detect(hisco_occupation, "Dentist") ~ "Dentist",
            TRUE ~ "Other"
        ),
        narrow_occupation = fct_relevel(narrow_occupation, "Other")
    )  %>% 
    filter(str_detect(birth_place, "Norge|Norway"),
    !str_detect(full_name, "Mannerbeck|Halvorsen"))

mapview(df_sf, zcol = "narrow_occupation", legend = TRUE, legend.pos = "bottomright", legend.title = "Birthplaces", legend.cex = 0.8, legend.mar = 5, legend.ncol = 1)
```

## Are engineers immigrants, or from very rural areas more than others?

- Just `3.6% of engineers were born outside of Sweden`, not significantly different from other occupations.

- Of Swedish born, priests come from places with low population density.

- No statistical differences in population density between the birthplaces of doctors, general managers, dentists, and engineers.

## Engineers are born in parishes with early access to electricity

![](assets/occupation_early_access.png){fig-align="center" width=100%}

## Where did engineers study?

![](assets/top_institutions.png){fig-align="center" width=100%}

<!-- ::: aside
Unsurpisingly, the Royal Institute of Technology (KTH) is the most common institution for engineers to study at, followed by Chalmers University of Technology.
::: -->

## Did engineers get advanced degrees?

![](assets/hisco_degree.png){fig-align="center" width=100%}

## Engineers in `electrical appliances and machinery` move the furthest from birthplace to study

@@

## Engineers in `electrical appliances and machinery` live further from their birthplace than other occupations

![](assets/distances_comparison.png){fig-align="center" width=80%}

::: aside
The median worker in Sweden in 1930 lived only 24km from their place of birth
:::

## Engineers live further from their birthplace than other occupations

![](assets/distance_moved_birth_to_current.png){fig-align="center" width=100%}

# Conclusion{#sec-conclusion background-color="#1C9E77"}

@@

# What to do next{#sec-conclusion background-color="#1C9E77"}

- Intergenerational mobility: did engineers' parents have similar jobs to doctors, dentists, businessmen?

### 2024-04-23 paper.qmd
change slide transition

---
title: "Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-who-scraper> and <https://github.com/j-jayes/svensk-industrikalender>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |
  This paper explores the transformative role of Artificial Intelligence in economic history research, through the examination of the skilled workforce in Sweden's electricity sector during the mid-20th century. By ingesting and analyzing two unique data sources—a set of biographical dictionaries and an industrial catalogue—the study investigates the origins and trajectories of high-skilled workers in electricity-related occupations in Sweden. Contrary to previous assumptions that local talent predominantly drove the technological advancements, the findings reveal a bifurcated labor market. Highly educated and skilled professionals, crucial in overseeing and advancing the electricity sector, frequently migrated for education and career opportunities, often traveling significant distances from their birthplaces. In contrast, local talent primarily occupied the burgeoning middle-skilled roles within the sector.
date: 12 January, 2024
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar  and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
# format:
#   docx:
#     toc: true
#     number-sections: true
#     highlight-style: github
execute: 
  echo: false
  warning: false
  message: false
bibliography: bibliography.bib
---

>>>>

---
title: "Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-who-scraper> and <https://github.com/j-jayes/svensk-industrikalender>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |
  This paper explores the transformative role of Artificial Intelligence in economic history research, through the examination of the skilled workforce in Sweden's electricity sector during the mid-20th century. By ingesting and analyzing two unique data sources—a set of biographical dictionaries and an industrial catalogue—the study investigates the origins and trajectories of high-skilled workers in electricity-related occupations in Sweden. Contrary to previous assumptions that local talent predominantly drove the technological advancements, the findings reveal a bifurcated labor market. Highly educated and skilled professionals, crucial in overseeing and advancing the electricity sector, frequently migrated for education and career opportunities, often traveling significant distances from their birthplaces. In contrast, local talent primarily occupied the burgeoning middle-skilled roles within the sector.

  New abstract.

  This paper investigates the origins and career trajectories of high-skilled workers in electricity-related occupations in Sweden during the mid-20th century. By ingesting and analyzing two unique data sources – a set of biographical dictionaries and an industrial catalogue – I ask where these workers came from, what and where they studied, and how international experience impacted their career paths.

  Jayes, Enflo and Molinder (2024) find that medium-skilled electricity-related jobs were filled by workers who were born near to these jobs. However, descriptive statistics about these high-skilled workers reveal a bifurcated labor market. Highly educated and skilled professionals, crucial in overseeing and advancing the electricity sector, frequently migrated for education and career opportunities, often traveling significant distances from their birthplaces to a small set of higher education institutions in Sweden and abroad. After their studies, Swedish engineers moved to opportunity, ending up on average ten times further from their birthplace than the median worker in 1930. Tentative findings on international experience show that engineers worked abroad at higher rates than comparable occupations, far more than doctors or dentists.

date: 12 January, 2024
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar  and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
# format:
#   docx:
#     toc: true
#     number-sections: true
#     highlight-style: github
execute: 
  echo: false
  warning: false
  message: false
bibliography: bibliography.bib
---
